---
title: "SEAHAWKS SHOULD HAVE RAN THE BALL"
subtitle: "Artificial American Football Coach with Decision Trees"
author: "Benjamin Jones"
date: "February 28, 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Have you ever dreamed of being an NFL coach? No? Me neither... 
Maybe you have a child on a pee wee football team in need of a coach? Also no? Same...
But perhaps you like data? If *you're* reading this, chances are you answered 'yes' to at least one of those questions.

The following dataset consists of football situations and the correct play to call for each situation. Let's see if it is possible to create a decision tree which, given a situation, can output the correct play to call. We'll try a few different tree models. The dataset is titled _Football Strategy_ and publicly available [\textcolor{blue}{here}](https://www.figure-eight.com/data-for-everyone/).

## Setup

Load the necessary libraries as well as authenticate the [\textcolor{blue}{data.world}](http://data.world) API:

```{r libraries, message=FALSE}
library(data.world)
library(jsonlite)
library(stringr)
library(rpart)
library(rpart.plot)
library(ipred)
library(randomForest)
library(rpart)
library(caret) 
setwd("~/Desktop")
dw_auth <- read_json("dw_cred.json")
saved_cfg <- save_config(dw_auth$token)
set_config(saved_cfg)
```

Query the data:

```{r query}
df <- query(qry_sql(paste0("SELECT football_scenarios_dfe_832307.antecedent as LABEL,
                            football_scenarios_dfe_832307.orig_antecedent as SITUATION ",
                            "FROM football_scenarios_dfe_832307")),
                            dataset="https://data.world/wppdatacatalogue/football-strategy")
```

\newpage
Preview the data:

```{r preview}
head(df, 3)
```

It's difficult to see, but an example of a full value of an entry in **SITUATION** is "It is first down and 10. The ball is on your opponent's 20 yardline. There is 3 seconds left in the second quarter. You are down by 3 points. Would you: "

## Cleaning

The entries of the football situation is a messy, as it is a long string. Fortunately, the values are consistent and we can do use patterns to pull out useful quantitative and qualitative features. For example:

"It is \textcolor{red}{first down} and \textcolor{cyan}{10}. The ball is on \textcolor{cyan}{your opponent's 20} yardline. There is \textcolor{cyan}{3} seconds left in the \textcolor{red}{second quarter}. You are \textcolor{cyan}{down by 3} points. Would you: "

"It is \textcolor{red}{third down} and \textcolor{cyan}{inches}. The ball is on \textcolor{cyan}{your 15} yardline. There is \textcolor{cyan}{3} seconds left in the \textcolor{red}{fourth quarter}. You are \textcolor{cyan}{up by 7} points. Would you: "

Numerical values used will be: yards to go, yard line, seconds left, and point differential. When distance is 'inches', it will be converted to 0.

Categorical valeus used will be: down and quarter.
### Remove Cases that Don't Follow the Pattern
```{r}
df <- df[grep("It is", df$SITUATION),]
df <- df[grep("The ball is", df$SITUATION),]
df <- df[grep("There is", df$SITUATION),]
```


### Down and Yards to Go
```{r down and ytg}
df$DOWN <- ""
df$YTG <- ""

for (i in 1:nrow(df)) {
ss1 <- str_split_fixed(df[i,2], " down ", 2)[1,1]
df$DOWN[i] <- str_split_fixed(ss1, "It is ", 2)[1,2]

ss2 <- str_split_fixed(df[i,2], " down ", 2)[1,2]
ss3 <- str_split_fixed(ss2, "and ", 2)
df$YTG[i] <- str_split_fixed(ss3, "\\.", 2)[2,1]
}

df$YTG <- str_replace(df$YTG, "inches", "0")
df$YTG <- as.numeric(df$YTG)
```

### Yardline
```{r position}
df$YL <- 0
for (i in 1:nrow(df)) {
ss1 <- str_split_fixed(df[i,2], "The ball is on ", 2)[1,2]
ss2 <- str_split_fixed(ss1, " yardline.", 2)[1,1]
df$YL[i] <- as.numeric(gsub("[^0-9.-]", "", ss2))
  if (grepl("your opponent", ss1)) {
    df$YL[i] <- 50 - df$YL[i]
  } else {
    df$YL[i] <- df$YL[i] - 50
  }
}
```

### Seconds and Quarter

```{r sec and q}
df$SEC <- 0
df$Q <- ""
for (i in 1:nrow(df)) {
  ss1 <- str_split_fixed(df[i,2], "There is ", 2)[1,2]
  if (grepl("minute", ss1)) {
    df$SEC[i] <- as.numeric(str_split_fixed(ss1, " minute", 2)[1,1])*60
  } else{
    df$SEC[i] <- as.numeric(str_split_fixed(ss1, " second", 2)[1,1])
  }
  ss2 <- str_split_fixed(ss1, " quarter", 2)[1,1]
  df$Q[i] <- str_split_fixed(ss2, " in the ", 2)[1,2]
}
```

### Score

```{r score}
df$PTS <-0
for (i in 1:nrow(df)) {
  ss1 <- str_split_fixed(df[i,2], "You are ", 2)[1,2]
  ss2 <- str_split_fixed(ss1, "by ", 2)[1,2]
  df$PTS[i] <- as.numeric(str_split_fixed(ss2, " points", 2)[1,1])
  if (grepl("down", ss1)) {
    df$PTS[i] <- df$PTS[i]*-1
  }
}
```

\newpage
### Final Dataset
Due to the fact there were only 9 cases of "Don't Know/it depends", and that it's not helpful, remove those cases.
```{r}
df$SITUATION <- NULL
df <- subset(df, df$LABEL != "Don't know / it depends")
df$LABEL <- as.factor(df$LABEL)
df$DOWN <- as.factor(df$DOWN)
df$Q <- as.factor(df$Q)
head(df, 5)
```

### Split into Training/Validation
While not shown here, distribution of labels between training and validation sets was assumed to be similar and was verified.

```{r train and val}
set.seed(74)
split_size <- floor(0.75*nrow(df))
split_index <- sample(seq_len(nrow(df)), size = split_size)

train <- df[split_index, ]
val <- df[-split_index, ]
```

\newpage
## Decision Trees

### Bootstrap Aggregation aka _Belicheck Yourself_
```{r bagging}
tree1 <- bagging(LABEL~., data=train)
results1 <- table(predict(tree1, val[,2:7]), val$LABEL)
confusionMatrix(results1)
#varImp(tree1)
```

\newpage
### Random Forest aka _Andy Feed_
```{r random forest}
tree2 <- randomForest(LABEL~., data=train)
results2 <- table(predict(tree2, val[,2:7]), val$LABEL)
confusionMatrix(results2)
#varImp(tree2)
```

\newpage
### Classification and Regression aka _Mike Shanahan-me-down_
```{r rpart}
tree3 <- rpart(LABEL~., data=train)
results3 <- table(predict(tree3, val[,2:7], type="class"), val$LABEL)
confusionMatrix(results3)
#varImp(tree3)
```

\newpage
## Conclusion
*Feature Importance* can also be evaluated. Similar to _Guess Who?_, feature importance describes the most efficient path of questioning to get to an answer. In this context, importance reveals which factors of the current play/game will lead to a play calling decision in the least amount of steps. In this experiment, the features ranked in descending order of importance are:

1. What down it is
2. Yards to Go
3. Yardline (position on field)
5. Score differential
6. Seconds remaining in the quarter
7. Which quarter it is



What I'm trying to say is... wait let me verify...

```{r seahawks}
i <- nrow(val)+1
val[i, 2] <- as.factor('second') #second down
val[i, 3] <- as.numeric(1) #1 yard to go
val[i, 4] <- as.numeric(99) #99 yard line
val[i, 5] <- as.numeric(26) #26s on the clock
val[i, 6] <- as.factor('fourth') #fourth quater
val[i, 7] <- as.numeric(-4) #down by 4

predict(tree1, val[i,2:7])
predict(tree2,val[i,2:7])
predict(tree3, val[i,2:7], type="class")
```

The Seahawks should have _kicked_ the ball? [\textcolor{blue}{Well nobody's perfect...}](https://www.youtube.com/watch?v=U7rPIg7ZNQ8)
